Modal operators for a constructive account of well quasi-orders

### 2025-02-26 slides-paris2025.tex
Backup slides

    \textbf{Lemma.} Let~$X$ be well\sinf. Let~$\alpha : \NN \to X$. Then there
    is an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq
    \ldots$.

    \emph{Proof.} Let~$K \defeq \{ n \in \NN \,|\, \neg \exists m > n\_
    \alpha\,n \leq \alpha\,m \}$ the set of indices of those terms which cannot
    appear as the first component of a good pair. If~$K$ is in bijection
    with~$\NN$, there is a subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq
    \ldots$ with~$k_0, k_1, \ldots \in K$. As~$X$ is well\sinf, this sequence
    is good, a contradiction.

@@

    \begin{column}{0.50\textwidth}
      \begin{block}{Key stability results}
        \justifying
        \only<1-5>{Assuming~\badbox{\textsc{lem}}
        and~\badbox{\textsc{dc}}}\only<6->{Constructively}, \ldots

@@

  \jnote{4}{
    Here is an example how the modal language helps us to give a constructive
    proof (for well\ind) of Dickson's lemma. In a similar vein Higman's lemma
    and Kruskal's theorem can be proven; with the exception of a preparatory
    first paragraph the original proofs can be copied word for word.

    \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
    is well.

    \emph{Lemma.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
    arbitrary topos. We need to show that~$\alpha$ is good. It suffices to
    prove that \emph{somewhere}, $\alpha$ is good, as goodness is a geometric
    implication (in fact even a geometric formula). Without loss of generality,
    we may suppose~\textsc{lem} as it holds \emph{somewhere}.

    Hence there is an infinite increasing subsequence
    \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
    As~$Y$ is well\ind, the sequence~$\gamma\,k_0, \gamma\,k_1, \ldots$
    is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
    also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
  }

@@

    \only<2>{
      \textbf{Examples for toposes.}
      \begin{itemize}
        \item $\mathrm{Set}$, the category of sets and maps.
        \item The category of sets and maps which are \hil{defined up to~$\boldsymbol\neg\neg$}.
        \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
        filter} of a \hil{forcing notion} (a quasi-order equipped with a
        coverage).
      \end{itemize}
      The following are \bad{not} toposes:
      \begin{itemize}
        \item The category of sets and partially defined maps.
        \item The category of abelian groups.
      \end{itemize}
    }

    \only<3->{
      \hil{Multiversal yoga:}
      \begin{enumerate}
        \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
        \smallskip
        \item A ring element is nilpotent iff all prime ideals \emph{everywhere} contain it.
        \smallskip
        \item For every inhabited set~$X$, \emph{proximally} there exists an
        enumeration~$\NN \twoheadrightarrow X$.
        \smallskip

@@

\begin{frame}{Toposophic constructive combinatorics}
  \justifying
  \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
  is well$\ind$.

  \emph{Proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
  arbitrary topos. We need to show that~$\alpha$ is good, so find indices~$n < m$ such that
  \[ \beta\,n \leq \beta\,m \quad\text{and}\quad \gamma\,n \leq \gamma\,m. \]

@@

  Thus there is an infinite increasing subsequence
  \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
  As~$Y$ is well\ind, the sequence~$(\gamma\,k_0, \gamma\,k_1, \ldots)$
  is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
  also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Toposophic constructive algebra}
  \justifying
  \textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.

  \emph{Classical proof.}
  \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
  The matrix~$M$ is surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
  is a contradiction to basic linear algebra. \qed
  \pause

  \emph{Multiversal constructive proof.}
  We may work \emph{somewhere} where \good{\textsc{lem}} holds. So \good{assume not}.
  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$.
  The matrix~$M$ is still surjective \emph{there}, and also over~$A/\mmm$. Since~$A/\mmm$ is a field, this
  is a contradiction to basic linear algebra. \qed
  \pause

@@

>>>>

    \textbf{Lemma.} Let~$X$ be well\sinf. Let~$\alpha : \NN \to X$. Then there
    is an infinite increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq
    \ldots$.

    \emph{Proof.} Let~$K \defeq \{ n \in \NN \,|\, \neg \exists m > n\_
    \alpha\,n \leq \alpha\,m \}$ be the set of indices of those terms which cannot
    appear as the first component of a good pair. If~$K$ is in bijection
    with~$\NN$, there is a subsequence~$\alpha\,k_0 \leq \alpha\,k_1 \leq
    \ldots$ with~$k_0, k_1, \ldots \in K$. As~$X$ is well\sinf, this sequence
    is good, a contradiction.

@@

    \begin{column}{0.50\textwidth}
      \begin{block}{Key stability results}
        \justifying
        \only<1-5>{Assuming~\badbox{\textsc{lem}}
        and~\badbox{\textsc{dc}}}\only<6->{\good{Constructively}}, \ldots

@@

@@

    \only<2>{
      \textbf{Examples for toposes.}
      \begin{itemize}
        \item $\mathrm{Set}$, the category of sets and maps.
        \item The category of sets and maps which are \hil{defined up to~$\boldsymbol{\neg\neg}$}.
        \item $\mathrm{Set}[G]$, the extension obtained by adding a \hil{generic
        filter} of a \hil{forcing notion} (a quasi-order equipped with a
        coverage).
      \end{itemize}
      The following are \bad{not} toposes:
      \begin{itemize}
        \item The category of sets and partially defined maps.
        \item The category of abelian groups.
      \end{itemize}
    }

    \visible<3->{
      \hil{Multiversal yoga:}
      \begin{enumerate}
        \item A quasiorder is well\ind iff \emph{everywhere}, every sequence is good.
        \smallskip
        \item A ring element is nilpotent iff \\ all prime ideals \emph{everywhere} contain it.
        \smallskip
        \item For every inhabited set~$X$, \\ \emph{proximally} there exists an
        enumeration~$\NN \twoheadrightarrow X$.
        \smallskip

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-4.68em}\includegraphics[width=1.2\paperwidth]{wqo-faded}\end{minipage}}
\begin{frame}{Multiversal constructive combinatorics}
  \justifying
  \textbf{Prop.} Let~$X$ and~$Y$ be well\ind quasi-orders. Then~$X \times Y$
  is well$\ind$.

  \emph{Multiversal constructive proof.} Let~$\alpha = (\beta,\gamma) : \NN \to X \times Y$ be a sequence in an
  arbitrary topos. We need to show that~$\alpha$ is good, i.\,e.\@ find indices~$n < m$ such that
  \[ \beta\,n \leq \beta\,m \quad\text{and}\quad \gamma\,n \leq \gamma\,m. \]

@@

  Thus there is an infinite increasing subsequence
  \[ \beta\,k_0 \leq \beta\,k_1 \leq \ldots. \]
  As~$Y$ is well\ind, the sequence~$(\gamma\,k_0, \gamma\,k_1, \ldots)$
  is good, so there exist~$i < j$ with~$\gamma\,k_i \leq \gamma\,k_j$. Since we
  also have~$\beta\,k_i \leq \beta\,k_j$, we are done. \qed
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\includegraphics[width=\paperwidth]{fr1}\end{minipage}}
\begin{frame}{Multiversal constructive algebra}
  \justifying
  \textbf{Thm.}
  Let~$M$ be a surjective matrix with more rows than columns over a
  ring~$A$. Then~$1 = 0$ in~$A$.
  \medskip

  \emph{Classical proof.}
  \bad{Assume not.} Then there is~a \bad{maximal ideal} $\mmm$.
  The matrix~$M$ is surjective over~$A/\mmm$. Since~$A/\mmm$ is a field, this
  is a contradiction to basic linear algebra. \qed
  \medskip
  \pause

  \emph{Multiversal constructive proof.}
  We may work \emph{somewhere} where \good{\textsc{lem}} holds. So \good{assume not}.
  \emph{Proximally,} there is a \good{maximal ideal}~$\mmm$.
  The matrix~$M$ is still surjective \emph{there}, and also over~$A/\mmm$. Since~$A/\mmm$ is a field, this
  is a contradiction to basic linear algebra. \qed
  \medskip
  \pause

@@

\backupstart

\section{Basics of forcing}

\begin{frame}{Ingredients for forcing}
  \vspace*{-0.3em}
  To construct a forcing extension, we require:
  \begin{enumerate}
    \item a base universe~$V$
    \item a preorder~$L$ of \hil{forcing conditions} in~$V$\!,
    pictured as \hil{finite approximations}

    (\emph{convention:} $\tau \preccurlyeq \sigma$ means that~$\tau$ is a
    better finite approximation than~$\sigma$)
    \item a \hil{covering system} governing how finite approximations evolve to
    better ones

    (for each~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
    P({\downarrow}\sigma)$, with a simulation condition)
  \end{enumerate}
  In the forcing extension~$V^\nabla$, there will then be a \hil{generic filter} (ideal
  object).
  \pause

  \vspace*{-1em}
  \begin{columns}
    \begin{column}{0.49\textwidth}
      \small
      \begin{block}{For the generic surjection~$\NN \twoheadrightarrow X$}
        \justifying
        Use \hil{finite lists}~$\sigma \in X^*$ as forcing conditions,
        where $\tau \preccurlyeq \sigma$ iff~$\sigma$ is an initial segment of~$\tau$,
        and be prepared to grow~$\sigma$ to \ldots
        \footnotesize
        \begin{enumerate}
          \item[(a)] one of~$\{ \sigma x \,|\, x \in X \}$, to make~$\sigma$ more defined
          \item[(b)] one of~$\{ \sigma \tau \,|\, \tau \in X^*, a \in
          \sigma\tau \}$, for any~$a \in X$, to make~$\sigma$ more surjective
        \end{enumerate}
        \vspace*{-0.4em}
      \end{block}
    \end{column}
    \pause

    \begin{column}{0.45\textwidth}
      \small
      \begin{block}{For the generic prime ideal of a ring~$A$}
        \justifying
        Use \hil{f.g.\@ ideals} as forcing conditions, where $\bbb \preccurlyeq
        \aaa$ iff~$\bbb \supseteq \aaa$, and be prepared to grow~$\aaa$ to \ldots
        \footnotesize
        \begin{enumerate}
          \item[(a)] one of~$\emptyset$, if~$1 \in \aaa$, to make~$\aaa$ more proper
          \item[(b)] one of~$\{ \aaa+(x), \aaa+(y) \}$, if~$xy \in \aaa$, to
          make~$\aaa$ more prime
        \end{enumerate}
        \vspace*{-0.4em}
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{-1cm}\includegraphics[width=\paperwidth]{forest-light-colored}\end{minipage}}
\begin{frame}{The eventually monad}
  Let~$L$ be a forcing notion.

  Let~$P$ be a monotone predicate on~$L$
  (if $\tau \preccurlyeq \sigma$, then $P\sigma \Rightarrow P\tau$). \\
  For instance, in the case~$L = X^*$:
  \begin{itemize}
    \item $\mathsf{Repeats}\, x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i = x_j$
    \item $\mathsf{Good}\, \,\,\,\,\,\,x_0\ldots x_{n-1} \defeqv \exists i\_ \exists j\_ i < j \wedge x_i \leq x_j$
    \quad (for some preorder~$\leq$ on~$X$)
  \end{itemize}
  \pause

  We then define~``\hil{$P \mid \sigma$}'' (``$P$ bars~$\sigma$'') inductively by the following clauses:
  \begin{enumerate}
    \item If~$P\sigma$, then~$P \mid \sigma$.
    \item If~$P \mid \tau$ for all~$\tau \in R$, where~$R$ is some covering
    of~$\sigma$, then~$P \mid \sigma$.
  \end{enumerate}
  So~$P \mid \sigma$ expresses in a \hil{direct inductive fashion}:
  \[ \text{``No matter
  how~$\sigma$ evolves to a better approximation~$\tau$, eventually~$P\tau$
  will hold.''} \]

  \pause
  We use quantifier-like notation: ``$\nabla(\tau \preccurlyeq \sigma)\_
  P\tau$'' means ``$P \mid \sigma$''.
\end{frame}}
% BOARD:
% - examples for P | σ
% - abuse of notation

\begin{frame}{Proof translations}
  \textbf{Thm.} Every~\textsc{iqc}-proof remains correct, with at most a
  polynomial increase in length, if throughout we
  replace
  \[\begin{array}{rcl@{\quad\text{where}\quad}rcl}
    \exists & \leadsto & \exists^\mathrm{cl},
    & \exists^\mathrm{cl} &\defeqv& \neg\neg\exists, \\
    \vee & \leadsto & \vee^\mathrm{cl},
    & \alpha \vee^\mathrm{cl} \beta &\defeqv& \neg\neg(\alpha \vee \beta), \\
    = & \leadsto & =^\mathrm{cl},
    & s =^\mathrm{cl} t &\defeqv& \neg\neg(s = t).
  \end{array} \]
  \pause

  \begin{columns}[c]
    \begin{column}{0.01\textwidth}
      \includegraphics[height=2.4em]{sheafification-man-2}
    \end{column}
    \quad
    \begin{column}{0.9\textwidth}
      \hil{When we say:}\ \ some statement ``holds in~$V^{\neg\neg}$'', \\
      \makesamewidth[l]{\hil{When we say:}}{\hil{we mean:}}\ \ its translation holds in~$V$.
    \end{column}
  \end{columns}
  \bigskip

  Similarly for arbitrary forcing extensions~$V^\nabla$, ``just with~$\nabla$
  instead of~$\neg\neg$''.
  \bigskip

  \pause
  \textbf{Ex.} As~$\neg\neg(\varphi \vee \neg\varphi)$ is a theorem
  of~\textsc{iqc}, the law of excluded middle holds in~$V^{\neg\neg}$.
\end{frame}

\newcommand{\defeqvi}{\quad iff\quad}
\begin{frame}{The $\nabla$-translation}
  \small
  \only<1>{For bounded first-order formulas over the (large) first-order signature which has
  \begin{enumerate}
    \scriptsize
    \item one sort~$\underline{X}$ for each set~$X$ in the base universe,
    \\[-1.2em]
    \item one~$n$-ary function symbol~$\underline{f} : \underline{X_1} \times
    \cdots \times \underline{X_n} \to \underline{Y}$ for each map~$f : X_1 \times
    \cdots \times X_n \to Y$,
    \\[-1.2em]
    \item one~$n$-ary relation symbol~$\underline{R} \hookrightarrow
    \underline{X_1} \times \cdots \times \underline{X_n}$ for each relation~$R
    \subseteq X_1 \times \cdots \times X_n$, and
    \\[-1.2em]
    \item an additional unary relation symbol~$G \hookrightarrow \underline{L}$
    (for the \emph{generic filter} of~$L$),
  \end{enumerate}
  we recursively define:}
  \scriptsize
  \only<2->{\vspace*{-0.4em}}
  \begin{tabbing}
    \quad \= $\sigma \forces \forces \forall(x\?\underline{X})\_ \varphi$ \=
    \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\
    \forall(x_0 \in X)\_ \tau \forces \varphi[\underline{x_0}/x]$.\qquad\quad \=
    $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
    \= $\sigma \forces \underline{R}(\underline{s_1},\ldots,\underline{s_n})$ \= \defeqvi $s = t$. \= \kill

    \> $\sigma \forces s = t$
    \> \defeqvi $\nabla \sigma\_ \llbracket s \rrbracket = \llbracket t \rrbracket$.
    \> $\sigma \forces \underline{R}(s_1,\ldots,s_n)$
    \> \defeqvi $\nabla\sigma\_ R(\llbracket s_1 \rrbracket,\ldots,\llbracket s_n \rrbracket)$. \\[0.3em]

    \> $\sigma \forces \varphi \Rightarrow \psi$
    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ (\tau \forces \varphi) \Rightarrow
    (\tau \forces \psi)$.
    \> $\sigma \forces G\tau$
    \> \defeqvi $\nabla\sigma\_ \sigma \preccurlyeq \llbracket\tau\rrbracket$. \\[0.3em]

    \> $\sigma \forces \top$ \> \defeqvi $\top$.
    \> $\sigma \forces \bot$ \> \defeqvi $\hil{$\nabla\sigma\_$}\ \bot$ \\[0.3em]

    \> $\sigma \forces \varphi \wedge \psi$
    \> \defeqvi $(\sigma \forces \varphi) \wedge (\sigma \forces \psi)$.
    \> $\sigma \forces \varphi \vee \psi$
    \> \defeqvi $\hil{$\nabla\sigma\_$}\ (\sigma \forces \varphi) \vee (\sigma \forces \psi)$. \\[0.3em]

    \> $\sigma \forces \forall(x\?\underline{X})\_ \varphi$
    \> \defeqvi $\textcolor{gray}{\forall(\tau \preccurlyeq \sigma)\_}\ \forall(x_0 \in X)\_ \tau \forces
    \varphi[\underline{x_0}/x]$.
    \> $\sigma \forces \exists(x\?\underline{X})\_ \varphi$
    \> \defeqvi $\hil{$\nabla\sigma\_$}\ \exists(x_0 \in X)\_ \sigma \forces \varphi[\underline{x_0}/x]$.
  \end{tabbing}
  \small
  \only<1>{Finally, we say that~$\varphi$ ``holds in~$V^\nabla$'' iff for all~$\sigma
  \in L$, $\sigma \forces \varphi$.}

  \footnotesize
  \begin{tabular}{@{}lp{0.27\textwidth}p{0.48\textwidth}@{}}
    \toprule
    forcing notion & statement about~$V^\nabla$ & external meaning \\
    \midrule
    surjection $\NN \twoheadrightarrow X$ &
    ``the gen.\@ surj.\@ is surjective'' &
    $\forall(\sigma{\in}X^*)\_ \forall(a{\in}X)\_ \nabla(\tau{\preccurlyeq}\sigma)\_ \exists(n{\in}\NN)\_ \tau[n] = a$. \\[0.4em]
    \pause

    map $\NN \to X$ &
    ``the gen.\@ sequence is good'' &
    $\mathsf{Good} \mid [\,]$. \\[0.4em]

    frame of opens &
    ``every complex number has a square root'' &
    For every open~$U \subseteq X$ and every cont.\@
    function $f : U \to \CC$, there is an open covering $U = \bigcup_i U_i$ such
    that for each index~$i$, there is a cont.\@ function $g : U_i \to \CC$
    such that~$g^2 = f$. \\[4.3em]

    big Zariski &
    ``$x \neq 0 \Rightarrow \text{$x$ inv.}$'' &
    If the only f.p.\@ $k$-algebra in which~$x = 0$ is the zero algebra,
    then~$x$ is invertible in~$k$.\\[1.5em]
    \pause

    little Zariski &
    ``every f.g. vector space does \emph{not not} have a basis'' &
    \makesamewidth[l]{}{\phantom{x}}\hil{Grothendieck's generic freeness lemma}
  \end{tabular}
\end{frame}

\begin{frame}{Outlook}
  \begin{block}{Passing to and from extensions}
    \justifying\small
    \textbf{Thm.} Let~$\varphi$ be a \hil{bounded first-order formula} not
    mentioning~$G$. In each of the following situations, we have that
    $\varphi$ holds in~$V^\nabla$ iff~$\varphi$
    holds in $V$:

    \vspace*{-0.5em}
    \begin{enumerate}
      \item $L$ and all coverings are inhabited (proximality). \\[-1em]
      \item $L$ contains a top element, every covering of the
      top element is inhabited, and~$\varphi$ is a coherent implication
      (positivity).
    \end{enumerate}
    \vspace*{-0.8em}
  \end{block}

  \vspace*{-1.5em}
  \begin{columns}
    \begin{column}{0.46\textwidth}
      \begin{block}{The mystery of nongeometric sequents}
        \justifying
        The \hil{generic ideal} of a ring is maximal:
        \vspace*{-1em}
        \[ (x \in \aaa \Rightarrow 1 \in \aaa) \Longrightarrow 1 \in \aaa + (x). \]

        The \hil{generic ring} is a field:
        \vspace*{-0.7em}
        \[ (x = 0 \Rightarrow 1 = 0) \Longrightarrow (\exists y\_ xy = 1). \]
        \vspace*{-1.4em}
      \end{block}

    \end{column}

    \begin{column}{0.50\textwidth}
      \begin{block}{Traveling the multiverse \ldots}
        \textsc{lem} is a \hil{switch} and \hil{holds positively};
        being countable is a \hil{button}.
        \medskip

        Every instance of \textsc{dc} \hil{holds proximally}.
        \medskip

        A geometric implication is provable iff it holds \hil{everywhere}.
        \vspace*{-0.2em}
      \end{block}
      \vspace*{-0.3em}
      \hfill\footnotesize\ldots{} upwards, but always keeping ties to the
      base.{\ }
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{More on forcing notions}
  \small
  \textbf{Def.} A \hil{forcing notion} consists of
  a preorder~$L$ of \hil{forcing conditions}, and
  for every~$\sigma \in L$, a set~$\Cov(\sigma) \subseteq
  P({\downarrow}\sigma)$ of \hil{coverings} of~$\sigma$
  such that: If~$\tau \preccurlyeq \sigma$ and~$R \in \Cov(\sigma)$, there
  should be a covering~$S \in \Cov(\tau)$ such that~$S \subseteq
  {\downarrow}R$.
  \bigskip

  {\centering\footnotesize\begin{tabular}{llll}
    \toprule
    & preorder~$L$ & coverings of an element~$\sigma \in L$ & filters of~$L$ \\
    \midrule
    \normalnumber{1} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$ & maps~$\NN \to X$ \\
    \normalnumber{2} & $X^*$ & $\{ \sigma x \,|\, x \in X \}$,\ \ $\{ \sigma\tau \,|\, \tau \in X^*, a \in \sigma\tau \}$ for each~$a \in X$ & surjections~$\NN \twoheadrightarrow X$ \\
    \normalnumber{3} & f.g. ideals & --- & ideals \\
    \normalnumber{4} & f.g. ideals & $\{ \sigma+(a), \sigma+(b) \}$ for each~$ab \in \sigma$,\ \ $\{\}$ if~$1 \in \sigma$ & prime ideals \\
    \normalnumber{5} & opens & $\mathcal{U}$ such that~$\sigma = \bigcup \mathcal{U}$ & points \\
    \normalnumber{6} & $\{\star\}$ & $\{ \star \,|\, \varphi \} \cup \{ \star \,|\, \neg\varphi \}$ &
    witnesses of~\textsc{lem}
    \\
    \bottomrule
  \end{tabular}\par}
  \bigskip

  \textbf{Def.} A \emph{filter} of a forcing notion~$(L,\mathrm{Cov})$
  is a subset~$F \subseteq L$ such that
  \vspace*{-0.4em}
  \begin{enumerate}
    \scriptsize
    \item $F$ is upward-closed: if~$\tau \preccurlyeq \sigma$ and if~$\tau \in F$, then~$\sigma \in F$; \\[-3.0em]
    \item $F$ is downward-directed: $F$ is inhabited, and if~$\alpha,\beta \in F$,
    then there is a common refinement~$\sigma \preccurlyeq \alpha,\beta$ such
    that~$\sigma \in F$; and \\[-2.0em]
    \item $F$ splits the covering system: if~$\sigma \in F$ and~$R \in
    \Cov(\sigma)$, then~$\tau \in F$ for some~$\tau \in R$.
  \end{enumerate}
\end{frame}

\backupend
